---
title: "R Notebook"
output: html_notebook
---

Immunoprofiling of adenocarcinomas of the pancreatobiliary system
=================================================================
Analysis of missing data & imputation
=================================================================

Import required packages
```{r}
# Build pre-requisites
#install.packages("plyr")
#BiocManager::install("impute") 


library(plyr)
library(impute)


```


Setting relative paths to import and export directories
```{r}

myWorkDirectory <- "~/tfcb-homework08/data/"
rawDataFileName <- paste(myWorkDirectory, "raw/pcbil_raw.csv", sep="")
imputedDataFileName <- paste(myWorkDirectory, "tidy/imputed_data.csv", sep="")


```
 

Importing raw dataset and changing "?" values to NA

```{r}

RawData <- read.csv(file = rawDataFileName, row.names = 39, colClasses= c(rep("numeric",38), rep("character",2)), na.strings = "?",quote="'" )
```

Creating working dataset for processing

```{r}
ProcessData <- RawData
```

Process column names 
```{r}
names(ProcessData) <- tolower(names(ProcessData)) #Converts column names to lowercase
names(ProcessData)[39] <- "clinical_diagnosis" #Names column 39 as 'clinical diagnosis'
#converts each element in the 'clinical_diagnosis' column to factors
ProcessData$clinical_diagnosis <- as.factor(ProcessData$clinical_diagnosis)

```



Rename the clinical diagnosis with the correct AJCC/IUCC-TNM 7 nomenclature

```{r}

revalue(ProcessData$clinical_diagnosis, c("Ampulla Ac" = "Ampullary carcinoma")) -> ProcessData$clinical_diagnosis
revalue(ProcessData$clinical_diagnosis, c("Distal Bile Duct Ac" = "Distal bile duct cancer")) -> ProcessData$clinical_diagnosis
revalue(ProcessData$clinical_diagnosis, c("Gallbladder Ac" = "Gallbladder cancer")) -> ProcessData$clinical_diagnosis
revalue(ProcessData$clinical_diagnosis, c("Hepatocellular Cancer" = "Hepatocellular carcinoma")) -> ProcessData$clinical_diagnosis
revalue(ProcessData$clinical_diagnosis, c("Intrahepatic Cholangiocarcinoma" = "Intrahepatic cholangiocarcinoma")) -> ProcessData$clinical_diagnosis
revalue(ProcessData$clinical_diagnosis, c("Pancreas Ac" = "Ductal pancreatic adenocarcinoma")) -> ProcessData$clinical_diagnosis
revalue(ProcessData$clinical_diagnosis, c("Perihilary Ac" = "Perihilar cholangiocarcinoma")) -> ProcessData$clinical_diagnosis
```


Fix labels and print a raw dataset summary
```{r}

#Fix one incorrect anatomical-based diagnosis (according to the authors data analysis)

ProcessData[rownames(ProcessData) == "Intrahepatic Cholangiocarcinoma|366" , "clinical_diagnosis"] <- "Gallbladder cancer"

#Lets observe the composition of the raw dataset by atanomical-based diagnosis

table(ProcessData$clinical_diagnosis)

```

Calculate the proportion of missing values (NA) for each immunohistochemical marker
```{r}
 

#"missingprop" fucntion takes a dataframe as an input, and returns a matrix containing the number of NA values, the length of the column and the proportion of NA entries 
  
missingprop <- function(dataframe){
  m <- sapply(dataframe, function(x){
    data.frame(nmiss=sum(is.na(x)), n=length(x), missingprop=round(sum(is.na(x))/length(x),2))
  })
  return(m)
}
  


#Takes the matrix output from calling the 'missingprop' function and transposes its rows and columns.
#Then it builds a data frame with each of the matrix's columns. Then it unlists each atomic component occurring in each columns (which are thus far lists).
#Finally, every column is coerced into 'na.probs' data frame 

na.probs<-as.data.frame(sapply(data.frame(t(missingprop(ProcessData))), unlist))
na.probs
```

Remove immunohistochemical markers with > 40% missing values:
```{r}

library("dplyr")
na.probs %>% filter(missingprop <0.4) %>%
#Reassigns back to 'na.probs' 
{.} -> na.probs

markers<-row.names(na.probs)


ProcessData %>% select(all_of(markers))%>% 
{.} -> ProcessData

```


Calculate the proportion of missing values (NA) for each cancer sample:

```{r}

#"missvalperrow" function takes a dataframe as an input, and returns a matrix containing the number of NA values per row, the length of the row and the proportion of NA entries per row

missvalperrow<-function(dataframe) {
  m2 <- apply(dataframe, 1, function(x){
    c(nmiss=sum(is.na(x)), n=length(x), missingprop=round(sum(is.na(x))/length(x),3))
  })
  return(m2)
}

#Takes the matrix output from calling the 'missvalperrow' function and transposes its rows and columns.
#Then it builds a data frame with each of the matrix's columns, and unlists each component occurring in each columns (which are thus far lists).
#Finally, every column is coerced into 'na.rows.prob' data frame 

na.rows.prob<-as.data.frame(sapply(data.frame(t(missvalperrow(ProcessData))), unlist))

```

Combine columns with the NA valued for each cancer sample, with the previously markers-trimmed data frame
```{r}

missrows<- data.frame(ProcessData,na.rows.prob)

```

Remove cancer samples with > 50% missing values:
```{r}


missrows %>% filter(missingprop <0.5) %>%
#Reassigns back to 'missrows' 
{.} -> missrows


#For imputation of missing immunohistochemical markers, remove columns without immunohistochemical markers info or/and numeric data types 
trimmedds<-missrows[1:26]

```


KNN-based imputation of remaining missing values
```{r}

#Impute NA entries
MatrixImputed <- impute.knn(as.matrix((trimmedds)))
DataImputed <- as.data.frame(MatrixImputed$data)

#Add back the 'clinical_diagnosis' column that was removed previous to imputation
DataImputed$clinical_diagnosis <- missrows$clinical_diagnosis
```

Save imputed data set in 'tidy' directory
```{r}

write.csv(DataImputed, imputedDataFileName, row.names=F, na="")

```
